---
title: "A brief introduction to the sf package"
output: 
  html_document:
    toc: true # table of content true
    toc_float: true
    toc_depth: 3  # upto three depths of headings (specified by #, ## and ###)
    number_sections: true  ## if you want number sections at each table header
    theme: united  # many options for theme, this one is my favorite.
    highlight: tango
    df_print: paged
---

## sf : the simple & tidy way of working with spatial data in R

The sf package is the (relatively) new kid on the block when it comes to working with spatial data in R. It provides a much more intuitive way and is easier to learn that the sp package. It provides a syntax and data-structures which are coherent with the tidyverse. A quick introduction can be found here:
https://mhweber.github.io/gis_in_action_r_spatial/2017/04/18/05-Spatial-Data-In-R-sf

It builds upon the 'simple features' standard (not R specific!) https://en.wikipedia.org/wiki/Simple_Features

Documentation on CRAN:
https://cran.r-project.org/web/packages/sf/sf.pdf

## Plotting a map : my workflow before using the sf-package

After reading in the shapefile you need to fortify it to be able to join the values which should be used to colour the map to it and plot it. Also the 'mapdata' dataset is created with the sole purpouse of generating the map : it contains a row for every edge-point which composes the polygons and to each point the value for the municipality mapped. 

```{r,echo=TRUE,message = TRUE,warnings=F}

library(pacman)
pacman::p_load(rgdal, rgeos)
# devtools::install_github("tidyverse/ggplot2")
library(ggplot2)

#read shapefile with rgdal (in past i've regularly used maptools::readShapePoly) this results in a nested datastructure (lists within lists) which is hard to decompose

shape <- readOGR("geodata/Shape_detailliert_SEEN_2016/UP_GEMEINDEN_SEEN_2016_F.shp")

#fortify -> generates a dataframe which ggplot2 can handle from a spatial object
shape2 <- fortify(shape, region='BFS')

# Dataset (Motorization and "generalabonnement"-owners)
mfz_ga <-readRDS("mfz_ga.rds")

#merge data with shapefile

mapdata <- merge(shape2, mfz_ga, by.x="id", by.y="bfs",all.x = TRUE, all.y = TRUE)

mapdata <- mapdata[order(mapdata$order),]

#how does the df look like?
mapdata[11300:11310,]

```

```{r,echo=TRUE,message = F,warnings=F}

# ggplot mapping  # data layer

ggplot(data=mapdata,aes(x=long, y=lat, group=group)) + 
  geom_path(color='gray') + 
  coord_equal()+ 
  geom_polygon(aes(fill=Anteil_GA),color = "white")+
  theme_void()+  
  theme(legend.key.size = unit(1,"line"),                                       
  #Grösse Legende
  legend.key.height= unit(0.5,"line"))+
  scale_fill_continuous(name="GA pro 100 EW")+
  labs(title="GA-Dichte im Kanton Zürich",
       subtitle="Anzahl Generalabonnemente pro 100 Einwohner",x="",y="")

```

## Plotting maps the 'tidy' way with the sf-package

```{r,echo=TRUE,message = F,warnings=F,include=TRUE}

library(pacman)

pacman::p_load(sf,dplyr,ggplot2)

#Shapefile (municipalities) - read_sf generates a tidy dataset, each row represents one municipality, the polygons are stored in a single special geo-variable, the 'geometry' column
gemeinden<- st_read('geodata/Shape_detailliert_SEEN_2016')

# Dataset (Motorization and GA-owners)
mfz_ga <-readRDS("mfz_ga.rds")

#join data - the dataset-structure stays the same, just 
gemdata <-gemeinden %>% left_join(mfz_ga, by=c("BFS"="bfs"))

#how does the df with class "sf" look like?
head(gemdata)

# dataframe with class sf can easily be ploted to get an overview! not so with sp...
plot(gemdata)

#map with ggplot2
sfmap <-ggplot()+
  geom_sf(data=gemdata,aes(fill=Anteil_GA),color = "white")+
  coord_sf(datum = NA)+  
  theme_void()+#Koordinatennetz verbergen
  theme(legend.key.size = unit(1,"line"),                                       
  legend.key.height= unit(0.5,"line"))+
  scale_fill_continuous(name="GA pro 100 EW")+
  labs(title="GA-Dichte im Kanton Zürich",
       subtitle="Anzahl Generalabonnemente pro 100 Einwohner",x="",y="")

sfmap

```

## sf dataframes: just a data.frame with a special geo-variable! We can use it like an ordinary structured dataframe.

```{r,echo=TRUE,message = F,warnings=F,cache=F}
library(gghighlight)
#library(devtools)
# devtools::install_github("statistikZH/statR")
# library(statR)

gghighlight_point(gemdata, aes(mfzpro100ew,Anteil_GA),Anteil_GA > 7 & mfzpro100ew<600 | 
                    mfzpro100ew>1000|
                    mfzpro100ew<700 & Anteil_GA<3|
                    mfzpro100ew>750 & Anteil_GA>9)+
  geom_point(aes(size=Anzahl_Einw/1000, color=Anteil_HTA))+
  # theme_stat()+
  scale_size(name="Einwohner")+
  labs(title="Motorisierungsgrad & Generalabonnements pro Gemeinde\n",
       y="Anzahl GA pro 100 Einwohner",x="Motorfahrzeuge pro 1000 Einwohner")

```

## This was just a foretaste...

Imagine we have several shapefiles which we want to plot layer by layer. With the old way we would need to fortify each shapefile...

```{r, eval=TRUE,include=TRUE}
# Shapefiles of the "Handlungsräume & Bauzonen"
handlungsraum<- st_read("geodata/handlungsraeume", stringsAsFactors = FALSE) 

bauzonen <- st_read("geodata/bauzonen", stringsAsFactors = FALSE) 

sfmap+
  geom_sf(data=handlungsraum,fill=NA)+
  geom_sf(data=bauzonen,alpha=0.5,color="white")+
  coord_sf(datum = NA)  
```


## geospatial operations: join etc!
```{r,eval=TRUE,include=TRUE}

# #CRS CH LV95 -> swiss projection
# st_crs(gemeinden)<- "+init=epsg:2056"

#Projektion wgs 84 lat lng
# sbb <- st_transform(sbb,"+init=epsg:4326 +proj=somerc +lat_0=46.95240555555556 +lon_0=7.439583333333333 +k_0=1 +x_0=2600000 +y_0=1200000 +ellps=bessel +units=m +no_defs")


```


